---
title: "Final Project"
author: "Chen Bistra 318322492 Eden Ahady 318948106 Lynn Molga Nagar 319090965"
output:
  pdf_document: default
---

# Introduction
NKI, consists of gene expression levels extracted from 272 tumors and is analyzed using about 1500 most varying genes
we are using pre-cleaned data

```{r, results='hide', message=FALSE, echo=FALSE}
library(plyr)
library(dplyr)
library(stringr)
library(gplots)
library(ggplot2)
library(forcats)
library(data.table)
library(reshape2)
library(affycoretools)
library(factoextra)
library(ggvis)
library(class)
library(gmodels)
library(plotly)
library(C50)
library(tidyverse)
library(skimr)     
library(ggthemes)
library(patchwork)  
library(GGally)     
library(corrplot)
library(psych)     
library(factoextra) 
library(regclass)
library(fmsb)
library(rpart.plot)
library(e1071)
library(plot.matrix)
library(gmodels)
library(plotly)
library(tidymodels)
library(rsample) 
library(ranger)
library(caret)
library(kknn)
```

```{r}
NKI_data<-read.csv("NKI_cleaned.csv")
```
this dataset has..
we have taken out Patient and ID as they are not features, in addition we have taken out eventdeath, survival and timerecurrece as they are the lables.


```{r}
print(paste("number of NAs in dataset:",sum(is.na(NKI_data))))
```

```{r}
# Histogram
ggplot(NKI_data, aes(x = age)) +
  geom_histogram(binwidth = 5, fill = "lightblue", color = "black") +
  labs(x = "Age", y = "Count") +
  ggtitle("Distribution of Age")

# Bar Plot
ggplot(NKI_data, aes(x = histtype)) +
  geom_bar(fill = "lightblue", color = "black") +
  labs(x = "Histological Type", y = "Count", title = "Histological Type Distribution")

```

```{r}
cleaned_data<-NKI_data[, c(3:4, 7:15)]

library(ggcorrplot)
cor_matrix <- cor(cleaned_data[-c(2)])
ggcorrplot(cor_matrix, type = "lower", lab = TRUE)

```
VIF
```{r}
X<-cleaned_data[,-(ncol(cleaned_data))]
print(paste("the VIF value of radius chemo is:",VIF(lm(chemo ~ .,data=X))))
print(paste("the VIF value of texture posnodes is:",VIF(lm(posnodes ~ .,data=X))))
```


```{r}
summary_df2 <- data.frame(
  Treatment = c("chemo", "hormonal", "amputation"),
  Alive = c(table(NKI_data$eventdeath[NKI_data$chemo == 1])[1],
            table(NKI_data$eventdeath[NKI_data$hormonal == 1])[1],
            table(NKI_data$eventdeath[NKI_data$amputation == 1])[1]),
  
  Dead = c(table(NKI_data$eventdeath[NKI_data$chemo == 1])[2],
           table(NKI_data$eventdeath[NKI_data$hormonal == 1])[2],
           table(NKI_data$eventdeath[NKI_data$amputation == 1])[2])
)

# Convert the Treatment column to a factor with desired order
summary_df2$Treatment <- factor(summary_df2$Treatment, levels = c("chemo", "hormonal", "amputation"))

# Reshape the data to a longer format
summary_df2_long <- tidyr::gather(summary_df2, Status, Count, Alive:Dead)

# Create the grouped bar plot
ggplot(summary_df2_long, aes(x = Treatment, y = Count, fill = Status)) +
  geom_col(position = "dodge") +
  scale_fill_manual(values = c("Alive" = "green", "Dead" = "red")) +
  labs(x = "Treatment", y = "Count", fill = "Status") +
  ggtitle("Alive and Dead Counts by Treatment")
```
```{r, }
# Calculate the average survival time for each grade
NKI_data$grade <- factor(NKI_data$grade)
avg_survival <- aggregate(survival ~ grade, data = NKI_data, FUN = mean)

ggplot(data = avg_survival, aes(x = grade, y = survival)) +
  geom_bar(stat = "identity", fill = "pink", width = 0.5) +
  labs(title = "Average Survival Time by Cancer Grade", x = "Cancer Grade", y = "Average Survival Time (Years)")
```

# **Normalization:**

We normalized by min-max and by Z-score, and wanted to check which one gave better results. 

### Normalize the data by MIN/MAX:

```{r}
# create the min/max normalization function:
normalize <- function(x) {
return ((x - min(x)) / (max(x) - min(x)))
}

#normalize the data without the classification column
minmax_data <- as.data.frame(lapply(cleaned_data, normalize))
str(minmax_data)
```
### Normalize the data by z-score:


```{r, message=FALSE, warning=FALSE}
# z-score normalization
zscore_data <- as.data.frame(scale(cleaned_data))
str(zscore_data)
```


```{r, message=FALSE, warning=FALSE, fig.show='hold', out.width="50%", out.height="50%"}

pca <- prcomp(minmax_data)


pca_to_show <- data.frame(
  PC1 = pca$x[, 1],
  PC2 = pca$x[, 2],
  classification = as.factor(cleaned_data$eventdeath)
)

ggplot(pca_to_show, aes(x = PC1, y = PC2, col = classification)) +
  geom_point()

#fviz_pca_var(pca,
#             col.var = "contrib", # Color by contributions to the PC
#             gradient.cols = c("#00AFBB", "#E7B800", "#FC4E07"),
#             repel = TRUE     # Avoid text overlapping
#             )

```
With Z-Score normalization:

```{r, message=FALSE, warning=FALSE, echo=FALSE, fig.show='hold', out.width="50%", out.height="50%"}

pca <- prcomp(zscore_data)
pca_to_show <- data.frame(
  PC1 = pca$x[, 1],
  PC2 = pca$x[, 2],
  classification = as.factor(cleaned_data$eventdeath)
)

ggplot(pca_to_show, aes(x = PC1, y = PC2, col = classification)) +
  geom_point()

#fviz_pca_var(pca,
#             col.var = "contrib", # Color by contributions to the PC
#             gradient.cols = c("#00AFBB", "#E7B800", "#FC4E07"),
#             repel = TRUE     # Avoid text overlapping
#             )
```

classification with tsne

```{r}
library(Rtsne)
tsne <- Rtsne(minmax_data,
  pca = FALSE, perplexity = 10,
  theta = 0.0
)

tsne_to_show <- data.frame(
  TSNE1 = tsne$Y[, 1],
  TSNE2 = tsne$Y[, 2],
  label=minmax_data$eventdeath,
  classification = minmax_data$eventdeath
)

ggplot(tsne_to_show, aes(
  x = TSNE1, y = TSNE2,
   label = label, col=classification
)) +
  geom_point() +
  ggrepel::geom_text_repel(cex = 2.5)
```


classification with umap
```{r}
library(uwot)
umap_result <- umap(minmax_data,
  n_neighbors = 15,
  min_dist = 1, spread = 5
)

umap_to_plot <- data.frame(
  UMAP1 = umap_result[, 1],
  UMAP2 = umap_result[, 2],
  label = minmax_data$eventdeath,
  classification = minmax_data$eventdeath
)

ggplot(umap_to_plot, aes(
  x = UMAP1, y = UMAP2,
  label = label, col = classification
)) +
  geom_point() +
  ggrepel::geom_text_repel(cex = 2.5)
```

```{r}
set.seed(123)  # For reproducibility
train_indices <- sample(1:nrow(NKI_data), 0.7 * nrow(NKI_data))  # 70% for training
train_data <- NKI_data[train_indices, ]
test_data <- NKI_data[-train_indices, ]

# Select the predictor variables
predictors <- c("age", "survival", "timerecurrence", "chemo", "hormonal", "amputation",
                "histtype", "diam", "posnodes", "grade", "angioinv", "lymphinfil")

# Train the k-NN model
k <- 11 # Set the number of neighbors
knn_model <- knn(train = train_data[, predictors],
                 test = test_data[, predictors],
                 cl = train_data$eventdeath,
                 k = k)
```

```{r}
knn_predictions <- knn_model

# Create a confusion matrix
confusion_matrix <- table(Actual = test_data$eventdeath, Predicted = knn_predictions)

# Print the confusion matrix
print(confusion_matrix)
```